Sample-Efficient Reinforcement Learning \\ with Implicitly Quantized Representations

### 2024-02-02 paper/main.tex
updated

\begin{abstract}
  Learning representations for reinforcement learning (RL) has shown much promise for continuous control. We propose an efficient representation learning method using only a self-supervised latent-state consistency loss. Our approach employs an encoder and a transition model to map observations to latent states and predict future states, respectively. We achieve high performance and prevent representation collapse by normalizing and bounding the latent representation such that rank in the representation is empirically preserved. Our method is straightforward, compatible with any model-free RL algorithm, and demonstrates state-of-the-art performance in continuous control benchmarks from DeepMind Control Suite.
\end{abstract}

@@

Recently, TD-MPC \citep{hansenTemporalDifferenceLearning2022} and TCRL \citep{zhaoSimplifiedTemporalConsistency2023}
have obtained state-of-the-art performance on continuous control benchmarks by learning
representations with self-supervised losses.
In particular, they leverage the latent-state consistency loss.
This requires an encoder $e_{\theta}$ which maps observations $o_{t}$ to latent states $z_{t}$ and a transition model $d_{\phi}$ which predicts
the next latent states $z_{t+1}$ given a latent state $z_{t}$ and an action $a_{t}$.
It then minimizes the distance between the next state predicted by the transition model $d_{\phi}(e_{\phi}(z_{t}), a_{t})$
and the next state predicted by the target encoder $e_{\bar{\phi}}(z_{t+1})$.
Self-supervised losses are susceptible to a problem known as representation collapse, where the encoder learns a constant
representation \cite{jingUnderstandingDimensionalCollapse2021}.
As such, it is common to combine the latent-state consistency loss with other loss terms, such as
minimizing the reward prediction error in the latent space
\citep{zhangLearningInvariantRepresentations2020,zhaoSimplifiedTemporalConsistency2023,hansenTemporalDifferenceLearning2022,geladaDeepMDPLearningContinuous2019,rezaei-shoshtariContinuousMDPHomomorphisms2022}.
This helps to prevent representation collapse at the cost of learning a task-specific representation.
% Typically it is combined with other loss terms such as reward or value prediction.

@@

\begin{algorithm}[tb]
   \caption{\our}
   \label{alg:main_alg}
   \renewcommand{\algorithmiccomment}[1]{\hfill\textcolor{gray}{\(\triangleright\) #1}}
\begin{algorithmic}
   \STATE {\bfseries Input:} Encoder $e_{\theta}$, transition $d_{\phi}$, critics $\{Q_{\psi_{1}}, Q_{\psi_{2}} \}$, policy $\pi_{\eta}$, learning rate $\alpha$, target network update rate $\tau$
   \FOR{$i$ {\bfseries to} $N_{\text{episodes}}$}
    \STATE Collect data in environment:
    \STATE $\qquad \mathcal{D} \leftarrow \mathcal{D} \cup \{o_{t}, a_{t}, o_{t+1}, r_{t+1}\}^{T}_{t=0}$
    \FOR{$i=1$ {\bfseries to} $T$}
        \STATE Update representation:
        \STATE $\qquad [\theta, \phi] \leftarrow [\theta, \phi] + \alpha \nabla \left( \mathcal{L}_{\text{rep}}(\theta, \phi; \mathcal{D}) \right)$  \COMMENT{\cref{eq:rep-loss}}
        \STATE Update critic
        \STATE \quad \quad $\psi \leftarrow \psi + \alpha \nabla \left( \mathcal{L}_{Q}(\psi; \mathcal{D}) \right)$ \COMMENT{\cref{eq:value-loss}}
        \IF{$i$ \% 2 == 0}
          \STATE Update actor less frequently than critic:
          \STATE $\quad \quad \eta \leftarrow \psi + \alpha \nabla \left( \mathcal{L}_{\pi}(\psi; \mathcal{D}) \right)$  \COMMENT{\cref{eq:policy-loss}}
          % \STATE $\bar{\eta} \leftarrow \tau \bar{\eta} + (1-\tau) \eta$
        \ENDIF
        \STATE Update target networks:
        \STATE $\quad \quad [\bar{\theta}, \bar{\phi}, \bar{\psi}] \leftarrow \tau [\bar{\theta}, \bar{\phi}, \bar{\psi}] + (1-\tau) [{\theta}, {\phi}, {\psi}]$
    \ENDFOR
   \ENDFOR
\end{algorithmic}
\end{algorithm}

@@

\textbf{Representation learning}
Our representation learning is based solely on the latent-state consistency loss,
%
\begin{align} \label{eq:rep-loss}
  \mathcal{L}_{\text{rep}}(\theta, \phi; \tau)
 % &= \E_{s_{t} \sim \mathcal{D}} \left[ \sum_{h=0}^{H-1} \| \hat{z}_{t+1} - z_{t+1} \|_{2}^{2} \right] \\
%&= \E_{(s_{t}, a_{t}, s_{t+1}) \sim \mathcal{D}}
% &= \E_{\tau \sim \mathcal{D}}
% \left[ \sum_{h=0}^{H-1} \gamma^{h} \| d_{\phi}(e_{\theta}(s_{t}), a_{t}) - e_{\bar{\theta}}(s_{t+1}) \|_{2}^{2} \right]
&= \sum_{h=0}^{H-1} \gamma^{h} \| d_{\phi}(\hat{z}_{t}, a_{t}) - e_{\bar{\theta}}(s_{t+1}) \|_{2}^{2},
\end{align}
%
which minimizes the distance between the next state predicted by the transition model $\hat{z}_{t+1} = d_{\phi}(\hat{z}_{t}, a_{t})$
and the next state predicted by the momentum encoder $\bar{z}_{t+1} = e_{\bar{\phi}}(s_{t+1})$.
The latent states are obtained with multi-step predictions in the latent space $\hat{z}_{t+1} = d(\hat{z}_{t}, a_{t})$
with $\hat{z}_{0} = e_{\theta}(s_{0})$.
The initial mapping to the latent space $\hat{z}_{0} = e_{\theta}(s_{0})$ uses the online encoder which
is being trained jointly with the transition model $d_{\phi}(\hat{z_{t}}, a_{t})$.
The target $e_{\bar{\phi}}(z_{t+1})$ is calculated with the momentum encoder which uses an exponential moving average (EMA)
of the encoder's weights $\bar{\theta} \leftarrow \tau \bar{\theta} + (1-\tau)\theta$.
% As such, gradient to not flow through the target network
The target network update rate is denoted $\tau$.

@@

\section{Conclusion}
\label{conclusion}
We have presented \our, a simple technique for learning representations which demonstrates state-of-the-art performance in
continuous control tasks.
Our normalization and bounding of the latent space empirically preserves the representation's matrix rank,
indicating that it alleviates representation collapse.
As such, \our learns a task agnostic representation and our experiments demonstrate that it shows
promise for incremental multi-task learning.
Importantly, our method is {\em (i)} straightforward, {\em (ii)} compatible with any model-free RL algorithm,
and {\em (iii)} learns a task-agnostic representation.
% Moreover, it can be combined with any model-free RL algorithm
% In our experiments we compared against both state-of-the-art model-free RL methods and state-of--the-art model-based RL methods.
% Our results demonstrate state-of-the-art based on our empirical results, we suggest further

>>>>

\begin{abstract}
  Learning representations for reinforcement learning (RL) has shown much promise for continuous control. We propose an efficient representation learning method using only a self-supervised latent-state consistency loss. Our approach employs an encoder and a dynamics model to map observations to latent states and predict future states, respectively. We achieve high performance and prevent representation collapse by normalizing and bounding the latent representation such that rank in the representation is empirically preserved. Our method is straightforward, compatible with any model-free RL algorithm, and demonstrates state-of-the-art performance in continuous control benchmarks from DeepMind Control Suite.
\end{abstract}

@@

Recently, TD-MPC \citep{hansenTemporalDifferenceLearning2022} and TCRL \citep{zhaoSimplifiedTemporalConsistency2023}
have obtained state-of-the-art performance on continuous control benchmarks by learning
representations with self-supervised losses.
In particular, they leverage the latent-state consistency loss.
This requires an encoder $e_{\theta}$ which maps observations $o_{t}$ to latent states $z_{t}$ and a dynamics model $d_{\phi}$ which predicts
the next latent states $z_{t+1}$ given a latent state $z_{t}$ and an action $a_{t}$.
It then minimizes the distance between the next state predicted by the dynamics model $d_{\phi}(e_{\phi}(z_{t}), a_{t})$
and the next state predicted by the target encoder $e_{\bar{\phi}}(z_{t+1})$.
Self-supervised losses are susceptible to a problem known as representation collapse, where the encoder learns a constant
representation \cite{jingUnderstandingDimensionalCollapse2021}.
As such, it is common to combine the latent-state consistency loss with other loss terms, such as
minimizing the reward prediction error in the latent space
\citep{zhangLearningInvariantRepresentations2020,zhaoSimplifiedTemporalConsistency2023,hansenTemporalDifferenceLearning2022,geladaDeepMDPLearningContinuous2019,rezaei-shoshtariContinuousMDPHomomorphisms2022}.
This helps to prevent representation collapse at the cost of learning a task-specific representation.\looseness-1
% Typically it is combined with other loss terms such as reward or value prediction.

@@

\begin{algorithm}[tb]
   \caption{\our}
   \label{alg:main_alg}
   \renewcommand{\algorithmiccomment}[1]{\hfill\textcolor{gray}{\(\triangleright\) #1}}
\begin{algorithmic}
   \STATE {\bfseries Input:} Encoder $e_{\theta}$, dynamics $d_{\phi}$, critics $\{Q_{\psi_{1}}, Q_{\psi_{2}} \}$, policy $\pi_{\eta}$, learning rate $\alpha$, target network update rate $\tau$
   \FOR{$i$ {\bfseries to} $N_{\text{episodes}}$}
    \STATE Collect data in environment:
    \STATE $\qquad \mathcal{D} \leftarrow \mathcal{D} \cup \{o_{t}, a_{t}, o_{t+1}, r_{t+1}\}^{T}_{t=0}$
    \FOR{$i=1$ {\bfseries to} $T$}
        \STATE Update representation:
        \STATE $\qquad [\theta, \phi] \leftarrow [\theta, \phi] + \alpha \nabla \left( \mathcal{L}_{\text{rep}}(\theta, \phi; \mathcal{D}) \right)$  \COMMENT{\cref{eq:rep-loss}}
        \STATE Update critic
        \STATE \quad \quad $\psi \leftarrow \psi + \alpha \nabla \left( \mathcal{L}_{Q}(\psi; \mathcal{D}) \right)$ \COMMENT{\cref{eq:value-loss}}
        \IF{$i$ \% 2 == 0}
          \STATE Update actor less frequently than critic:
          \STATE $\quad \quad \eta \leftarrow \psi + \alpha \nabla \left( \mathcal{L}_{\pi}(\psi; \mathcal{D}) \right)$  \COMMENT{\cref{eq:policy-loss}}
          % \STATE $\bar{\eta} \leftarrow \tau \bar{\eta} + (1-\tau) \eta$
        \ENDIF
        \STATE Update target networks:
        \STATE $\quad \quad [\bar{\theta}, \bar{\phi}, \bar{\psi}] \leftarrow \tau [\bar{\theta}, \bar{\phi}, \bar{\psi}] + (1-\tau) [{\theta}, {\phi}, {\psi}]$
    \ENDFOR
   \ENDFOR
\end{algorithmic}
\end{algorithm}

@@

\textbf{Representation learning}
Our representation learning is based solely on the latent-state consistency loss,
%
\begin{align} \label{eq:rep-loss}
  \mathcal{L}_{\text{rep}}(\theta, \phi; \tau)
 % &= \E_{s_{t} \sim \mathcal{D}} \left[ \sum_{h=0}^{H-1} \| \hat{z}_{t+1} - z_{t+1} \|_{2}^{2} \right] \\
%&= \E_{(s_{t}, a_{t}, s_{t+1}) \sim \mathcal{D}}
% &= \E_{\tau \sim \mathcal{D}}
% \left[ \sum_{h=0}^{H-1} \gamma^{h} \| d_{\phi}(e_{\theta}(s_{t}), a_{t}) - e_{\bar{\theta}}(s_{t+1}) \|_{2}^{2} \right]
&= \sum_{h=0}^{H-1} \gamma^{h} \| d_{\phi}(\hat{z}_{t}, a_{t}) - e_{\bar{\theta}}(s_{t+1}) \|_{2}^{2},
\end{align}
%
which minimizes the distance between the next state predicted by the dynamics model $\hat{z}_{t+1} = d_{\phi}(\hat{z}_{t}, a_{t})$
and the next state predicted by the momentum encoder $\bar{z}_{t+1} = e_{\bar{\phi}}(s_{t+1})$.
The latent states are obtained with multi-step predictions in the latent space $\hat{z}_{t+1} = d(\hat{z}_{t}, a_{t})$
with $\hat{z}_{0} = e_{\theta}(s_{0})$.
The initial mapping to the latent space $\hat{z}_{0} = e_{\theta}(s_{0})$ uses the online encoder which
is being trained jointly with the dynamics model $d_{\phi}(\hat{z_{t}}, a_{t})$.
The target $e_{\bar{\phi}}(z_{t+1})$ is calculated with the momentum encoder which uses an exponential moving average (EMA)
of the encoder's weights $\bar{\theta} \leftarrow \tau \bar{\theta} + (1-\tau)\theta$.
% As such, gradient to not flow through the target network
The target network update rate is denoted $\tau$.

@@

\section{Conclusion}
\label{conclusion}
We have presented \our, a simple technique for learning representations which demonstrates state-of-the-art performance in
continuous control tasks.
Our normalization and bounding of the latent space empirically preserves the representation's matrix rank,
indicating that it alleviates representation collapse.
As such, \our learns a task-agnostic representation and our experiments demonstrate that it shows
promise for incremental multi-task learning.
Importantly, our method is {\em (i)} straightforward, {\em (ii)} compatible with any model-free RL algorithm,
and {\em (iii)} learns a task-agnostic representation.
% We hope that our representation learning technique is
% Moreover, it can be combined with any model-free RL algorithm
% In our experiments we compared against both state-of-the-art model-free RL methods and state-of--the-art model-based RL methods.
% Our results demonstrate state-of-the-art based on our empirical results, we suggest further
